<!--
%\VignetteIndexEntry{01.3 New and relevant packages}
%\VignettePackage{UseBioconductor}
%\VignetteEngine{knitr::knitr}
-->

```{r setup, echo=FALSE}
library(UseBioconductor)
stopifnot(BiocInstaller::biocVersion() == "3.1")
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# New and relevant packages

Martin Morgan<br/>
February 3, 2015

## _Bioconductor_ packages

`r Biocpkg("GenomicFiles")`

- `GenomicFiles()` to coordinate on-disk files with
  regions-of-intereset
- `reduceBy*()` to iterate over regions or files
- `reduceByYield()` to iterate through files, e.g., `BamFile(...,
  yieldSize=1000000)`

`r Biocpkg("GenomeInfoDb")`

- Manipulate `seqlevels()`, e.g., keep only standard chromsomes
  (`keepStandardChromosomes()`), place in standard order
  (`sortSeqlevels()`), rename (`renameSeqlevels()`)

`r Biocpkg("BiocStyle")`

- Standardize formatting
- HTML vignettes _much_ more usable than PDF
- Supports `rmarkdown`

(`Rhtslib`)

## _Bioconductor_ functionality

`GenomicAlignments::readGAlignmentPairs()` / `readGAlignmentsList()`

`Rsamtools::pileup()`

`VariantAnnotation::VRanges()`

`findOverlaps()`: significant performance improvements

`AnnotationDbi::mapIds()`

## Contributed packages

`r Biocpkg("csaw")`

- Quantitative variation in ChIP binding; differential binding between
  treatment groups more relevant binding _per se_.
- See the _Bioconductor_
  [newsletter](http://bioconductor.org/help/newsletters/2015_January/#overview-of-the-csaw-package),
  especially final question for statistical rationale and advanced
  implementation (very interesting reading).

`r Biocpkg("Rariants")`

- SNVs from comparative shifts in non-consensus base call frequencies.

`r Biocpkg("SGSeq")`

- Prediction, quantification and visualization of alternative
  transcript events (rather than assessment of whole-transcript
  abundance).

`r Biocpkg("ballgown")` 

- differential expression of assembled transcriptomes, 

`r Biocpkg("derfinder")` 

- base-level differential expression

`r Biocpkg("AnalysisPageServer")`

- interactive reports(!)

`r Biocpkg("ggtree")`

- Impressive visualization of phylogenetic trees

## Etc.

- [New packages](http://bioconductor.org/developers/new_packages/)
  summary (also posted to twitter)
